---
title: "Tidyverse carpentry - day 1, afternoon"
author: "Desiree Treichler"
date: "6 juni 2017"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Preparations: load data

```{r, message = FALSE, warning = FALSE, error = FALSE}
library(tidyverse)
gapminder <- read_csv("../Data/gapminder-FiveYearData.csv")
```


# The pipe: %>%  
We could write
```{r, message = FALSE, warning = FALSE, error = FALSE}
rep("This is an example", times=3)
```
or, do the same with a pipe:
```{r, message = FALSE, warning = FALSE, error = FALSE}
"This is an example" %>% rep(times=3)
```

##Make a new dataframe and show the head
```{r, message = FALSE, warning = FALSE, error = FALSE}
year_country_gdp <- select(gapminder,year,country,gdpPercap)
knitr::kable(head(year_country_gdp)) # kable from the knitr package formats it nicely in markdown
```


-- or, do the same with a pipe:
```{r, message = FALSE, warning = FALSE, error = FALSE}
year_country_gdp <- gapminder %>%
  select(year,country, gdpPercap)
knitr::kable(head(year_country_gdp))
```


We can also do a filter and plot directly.

```{r, message = FALSE, warning = FALSE, error = FALSE}
gapminder %>% filter(year==2002) %>% 
  ggplot(mapping=aes(x=continent, y=pop)) +
  geom_boxplot()
```

Or filter and select in the same command. 
```{r, message = FALSE, warning = FALSE, error = FALSE}
year_country_gdp_euro <- gapminder %>% 
  filter(continent=="Europe") %>% 
  select(year,country,gdpPercap)
knitr::kable(head(year_country_gdp_euro))
```

**Challenge 1: A single command that produces a data frame that has the Norwegian values for gdpPercap, liofeExp and year. How many rows does your data frame have and why?**

```{r, message = FALSE, warning = FALSE, error = FALSE}
Norway <- gapminder %>%
  filter(country=="Norway") %>%
  select(gdpPercap, lifeExp, year)
Norway
```
Answer: the dataset has twelve rows. 


## Grouping  
Take the gapminder dataset, group it by continent, and summarize per group.
```{r, message = FALSE, warning = FALSE, error = FALSE}
gapminder %>% 
  group_by(continent) %>%
  summarize(mean_gdpPercap=mean(gdpPercap))
```

Add some plotting to the pipe.
```{r, message = FALSE, warning = FALSE, error = FALSE}
gapminder %>% 
  group_by(continent) %>%
  summarize(mean_gdpPercap=mean(gdpPercap)) %>%
  ggplot(mapping = aes(x=continent, y = mean_gdpPercap))+
  geom_point()
```

**Challenge 2: Calculate the average life expectancy per country in Asia. Which one has the longest onem which one the shortest?**

```{r, message = FALSE, warning = FALSE, error = FALSE}
gapminder %>%
  filter(continent=="Asia") %>%
  group_by(country) %>%
  summarize(mean_lifeExp = mean(lifeExp)) %>%
  filter(mean_lifeExp==min(mean_lifeExp)|mean_lifeExp==max(mean_lifeExp))
```
The last filter acts as a selector, and choses only the min/max value. 

Alternatively, use arrange to order the list:
```{r, message = FALSE, warning = FALSE, error = FALSE}
gapminder %>%
  filter(continent=="Asia") %>%
  group_by(country) %>%
  summarize(mean_lifeExp = mean(lifeExp)) %>%
  arrange(mean_lifeExp)
```

Or, we could solve this graphically:
```{r, message = FALSE, warning = FALSE, error = FALSE}
gapminder %>%
  filter(continent=="Asia") %>%
  group_by(country) %>%
  summarize(mean_lifeExp = mean(lifeExp)) %>%
  ggplot(mapping = aes(x=reorder(country,mean_lifeExp), y = mean_lifeExp))+ geom_point() + 
  coord_flip()
```
  
To sort the y axis, specify the sorting variable directly in the aes() function.


## Mutate
Mutate adds new variables and preserves existing; transmute drops existing variables.
Do some mutating and pipe the result directly to the head function:
```{r, message = FALSE, warning = FALSE, error = FALSE}
gapminder %>%
  mutate(gdp_billion=gdpPercap*pop/10^9) %>%
  knitr::kable(head())
```
This shows the GDP in billions for every country and every year.

We can also summarize this further, and group by continent/year.

```{r, message = FALSE, warning = FALSE, error = FALSE}
gapminder %>%
  mutate(gdp_billion=gdpPercap*pop/10^9) %>%
group_by(continent,year) %>%
  summarize(mean_gdp_billion=mean(gdp_billion))
```

We can create a summary dataset for each country to use later.
```{r, message = FALSE, warning = FALSE, error = FALSE}
gapminder_country_summary <- 
  gapminder %>%
  group_by(country) %>%
  summarize(mean_lifeExp = mean(na.omit(lifeExp))) %>% #na.omit would remove NaNs, if there are any
  arrange(mean_lifeExp)
```


## Maps
Now, let's add some maps! First, load the package. 
```{r, message = FALSE, warning = FALSE, error = FALSE}
library(maps)
```

Then, get some coordinates for the entire world. 
```{r, message = FALSE, warning = FALSE, error = FALSE}
map_data("world") %>%
  knitr::kable(head())
```

## rename and left join
Unfortunately, in this dataset, the countries are called "region".  Rename this, join it with the country summary gapminder dataset, and send it directly to the plot. 
```{r, message = FALSE, warning = FALSE, error = FALSE}
map_data("world") %>%
  rename(country=region) %>%
  left_join(gapminder_country_summary, by="country") %>%
  ggplot() + geom_polygon(aes(x=long, y=lat, group=group, fill=mean_lifeExp)) +
  scale_fill_gradient(low="blue", high="red") # what colours to use
```
The holes are where the gapminder country named don't match the names of the map_data region names. 