---
title: "tidyr.R"
author: "Desiree Treichler"
date: "6 juni 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Download some new data - first, try automatically.

```{r, message = FALSE, warning = FALSE, error = FALSE}
download.file(url = "http://docs.google.com/spreadsheet/pub?key=phAwcNAVuyj0NpF2PTov2Cw&output=xlsx", 
              destfile = "../Data/indicator gapminder infant_mortality.xlsx")

download.file(url = "http://docs.google.com/spreadsheet/pub?key=phAwcNAVuyj0NpF2PTov2Cw&output=xlsx", 
              destfile = "../Data/indicator undata total_fertility.xlsx")
```

To read them, we need the readxl library. 
```{r, message = TRUE, warning = TRUE, error = TRUE}
library("readxl")
library("tidyverse")
raw_fert <- read_excel(path="../Data/indicator undata total_fertility.xlsx", sheet="Data")
raw_infantMort <- read_excel(path="../Data/indicator gapminder infant_mortality.xlsx",sheet="Data")
```

Something is wrong with the files, so we download them manually as ...2.xlsx.
Read in and preview!
```{r, message = FALSE, warning = FALSE, error = FALSE}
raw_fert <- read_excel(path="../Data/indicator undata total_fertility2.xlsx", sheet="Data")
raw_infantMort <- read_excel(path="../Data/indicator gapminder infant_mortality2.xlsx",sheet="Data")
```

```{r}
knitr::kable(head(raw_fert))
```
```{r, message = FALSE, warning = FALSE, error = FALSE}
knitr::kable(head(raw_infantMort))
```


## Tidy up your data!
Typically, we want one observation per row, and several attributes in the columns. We want to get from the long/wide formato to the tidy format that suits R best. 
Long: everything stacked on top of each other.   
Wide: as the input above -- a classical table, e.g. first column the ID, second to last column data for different years -> this we could tidy up: one column for the year, one column for the measurement.

The gather function can do this:

```{r, message = FALSE, warning = FALSE, error = FALSE}
fert <- raw_fert %>%
  rename(country=`Total fertility rate`) %>%  # so that it fits to our country names!
  gather(key=year,value=fert,-country) %>% # gather everything but the first column
  mutate(year=as.integer(year))
knitr::kable(head(fert))
```

Note that the variable "Total fertility rate" needs single backticks - as seen in the preview!

